---
title: "Advent of Code: Day 5"
author: "Keyi Long"
date: "2/19/2021"
output: pdf_document
---

# Reference  
RegEx Cheatsheet: https://rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf    
Stringr Cheatsheet: https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_strings.pdf    
Strings: https://jrnold.github.io/r4ds-exercise-solutions/strings.html

```{r}
library(dplyr)
library(stringr)

# Read in dataset
raw <- read.csv("./input/input5.txt", header = F)
```

# Functions
## Detect Patterns.   
* **str_detect(string, pattern = "")**, equivalent to `grepl(pattern, x)`.    
  
## Locate Patterns.    
* **str_count(string, pattern = "")**, or return the length of all matches generated from `gregexpr()` or `stringr::str_locate_all()`   

# Patterns
* **`\\1`** pattern is called a backreference. It matches whatever the first group matched.   
* **{n}** indicates exactly n matches. We use **`\\1`** instead of **`{2}`** to indicate **twice in a row**.   
* Can use **.** to represent any character, or use "[a-z]" or [[:alpha:]].    
* *: matches at least 0 times   

## Some examples    
```{r}
grepl("([a-z]){2}", "abccd")
grepl("([a-z])\\1", "abccd")

grepl("([a-z]){2}", "abcdc")
grepl("([a-z])\\1", "abcdc")
```

**`(.).\\1.\\1`**: A character followed by any character, the original character, any other character, the original character again. E.g. "abaca"    
**`(..)\\1`**: Any two characters repeated. E.g. "a1a1", "chch"    
**`(.).*\\1`**: A character repeated, with zero or more characters between them. E.g. "aa", "aba", "abca"       


# Part 1    
Use 3 flags to indicate whether the given strings meet the corresponding criterion:   
* Flag 1: contains at least three vowels.    
* Flag 2: contains at least one letter that appears twice in a row.    
* Flag 3: does not contain the strings `ab`, `cd`, `pq`, or `xy`. 

```{r}
data1 <- raw %>% 
  # Flag 1: contains at least three vowels
  mutate(flag1 = (str_count(V1, "[aeiou]") >= 3)) %>% 
  # Flag 2: contains at least one letter that appears twice IN A ROW
  mutate(flag2 = str_detect(V1, "(.)\\1")) %>% 
  # Flag 3: does NOT contain the strings ab, cd, pq, or xy
  mutate(flag3 = !str_detect(V1, "ab|cd|pq|xy")) %>% 
  filter(flag1*flag2*flag3 == TRUE)

nrow(data1)
```

```{r}
data2 <- raw %>% 
  # Flag 1: contains at least three vowels
  rowwise() %>% 
  # gregexpr: find starting position and length of all matches
  mutate(flag1 = (length(gregexpr('[aeiou]', V1)[[1]]) >= 3)) %>% 
  # Flag 2: contains at least one letter that appears twice IN A ROW
  mutate(flag2 = grepl("([a-z])\\1", V1)) %>% 
  # Flag 3: does NOT contain the strings ab, cd, pq, or xy
  mutate(flag3 = !grepl("ab|cd|pq|xy", V1)) %>% 
  filter(flag1*flag2*flag3 == TRUE)

nrow(data2)
```

# Part 2    
Use 2 flags to indicate whether the given strings meet the corresponding criterion:   
* Flag 1: contains a pair of any two letters that appears at least twice without overlapping.   
* Flag 2: contains at least one letter which repeats with exactly one letter between them  
```{r}
data3 <- raw %>% 
  # Flag 1: contains a pair of any two letters that appears at least twice without overlapping
  mutate(flag1 = str_detect(V1, "(..).*\\1")) %>% 
  # Flag 2: contains at least one letter which repeats with exactly one letter between them
  mutate(flag2 = str_detect(V1, "(.).\\1")) %>% 
  filter(flag1 == TRUE & flag2 == TRUE)

nrow(data3)
```

When using `grepl`, there is an issue when the repeated pair of characters are not the first two characters. But we do not need to worry about this when using `str_detect`.    
```{r}
grepl("(..).*\\1", "qwabxyzab") # 'ab' repeated but cannot be detected
str_detect("qwabxyzab", "(..).*\\1") # 'ab' repeated and detected
grepl(".*(..).*\\1", "qwabxyzab") # 'ab' detected
```

```{r}
data4 <- raw %>% 
  # Flag 1: contains a pair of any two letters that appears at least twice without overlapping
  mutate(flag1 = grepl(".*(..).*\\1", V1)) %>% 
  # Flag 2: contains at least one letter which repeats with exactly one letter between them
  mutate(flag2 = grepl("(.).\\1", V1)) %>% 
  filter(flag1 == TRUE & flag2 == TRUE)

nrow(data4)
```



```{r, include = FALSE}
# Part 1
# Flag 1: at least three vowels
gregexpr('[aeiou]', "ugknbfddgicrmopn")[[1]] # find starting position and length of all matches
length(gregexpr('[aeiou]', "ugknbfddgicrmopn")[[1]])

regexpr('[aeiou]', "wugknbfddgicrmopn")[[1]] # find starting position and length of first match

# Flag 2: at least one letter that appears twice in a row
grepl("(.){2}", "abc11d")
grepl("(.)\\1", "abc11d")

grepl("([a-z]){2}", "abcdc")
grepl("([a-z])\\1", "abcdc")

str_detect("aasaba", '([a-z])\\1')

# Flag 3: does not contain the strings ab, cd, pq, or xy
grepl("ab|cd|pq|xy", "acbusahbkdasbxzy")
str_detect("aspqnsxyen", "ab|cd|pq|xy")


# Part 2
# Flag 1: contains a pair of any two letters that appears at least twice without overlapping
str_detect("qwabxyzab", "(..).*\\1")
grepl("(..).*\\1", "qwabxyzab") 
grepl(".*(..).*\\1", "qwabxyzab") #ab

# Flag 2: contains at least one letter which repeats with exactly one letter between them
str_detect("ieodomkazucvgmuy", "(.).\\1")
str_detect("ieodomkazucvgmuy", "([a-z])[a-z]\\1")
str_detect("oeodaomkazgm", "([a-z])([a-z])\\1")

data5 <- raw %>% 
  # Flag 1: contains a pair of any two letters that appears at least twice without overlapping
  rowwise() %>% 
  mutate(flag1 = ifelse(regexpr(".*(..).*\\1", V1)[[1]] >= 1, TRUE, FALSE)) %>% 
  # Flag 2: contains at least one letter which repeats with exactly one letter between them
  mutate(flag2 = ifelse(regexpr("(.).\\1", V1)[[1]] >= 1, TRUE, FALSE)) %>% 
  filter(flag1 == TRUE & flag2 == TRUE)

nrow(data5)
```

